Two years of CiTO annotations

### 2022-11-13 article.Rmd
Filled in some gaps.

[//]: <> (|Research article|European Registry of Materials: global, unique identifiers for (undisclosed) nanomaterials [@citesForInformation:vanRijn2022]||| --> THIRD YEAR!)

@@

Markdown template...

To make life easier for authors, and following a Twitter discussion in Spring 2021, we developed a Markdown template with native CiTO support. Here, the author merely indicates the CiTO type when they cite the article. This is using a method introduced by Krewinkel et al. [@usesMethodIn:Krewinkel2017].

The manuscript can then easily be saved as a Word file to enable submission to the journal. The *Journal of Cheminformatics* template is available from our GitHub repository, and authors and editors should feel free to adapt it to their own journal needs.

@@

Formatting issues

Citation level intention annotation: only Markdown

@@

With X articles published in the CiTO Collection, the adoption is not as high as one would have hoped.
Nevertheless, the pilot is triggering interest with both authors and publishing platforms. The Scholia
use case shows how the information can be used downstream. Further uptake of this idea of typed citations
depends on the combined willingness of journal editors, authors, publishers and indexing services alike.
The rise of services like scite.ai shows that the research community needs this kind of information.

#### Availability of data and materials
Text for this section.

@@

#### Funding
Text for this section.

>>>>

@@

To make life easier for authors, and following a Twitter discussion in Spring 2021, a Markdown template was developed with native CiTO support: https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/markdown.html. Here, the author merely indicates the CiTO type when they cite the article. This is using a method introduced by Krewinkel et al. [@usesMethodIn:Krewinkel2017].
The manuscript can then be converted to a Microsoft Word file with Pandoc (https://pandoc.org/) for submission to the journal.

The *Journal of Cheminformatics* template is available from the journal's GitHub organization, and authors and editors should feel free to adapt it to the needs of other journals. The BioHackrXiv (https://biohackrxiv.org/) preprint server also support CiTO annotations [@cito:citesAsDataSource:Prins2021] and this template can be found at https://github.com/biohackrxiv/publication-template.

@@

Because many publishing platforms currently do not support display of citation level intention annotation,
the default model only provides the annotation in the bibliography. The choice was made to use the `bibnotes`
idea which allows giving additional notes to a reference in the bibliography. The CiTO Pilot uses
the *safe CURIE* standard proposed by the W3C in 2010 [@usesMethodIn:BirBeck2010]. The Markdown templates
implement this idea. However, typesetting in the *Journal of Cheminformatics* has shown to be inconsisten,
even within single articles, suggesting additional training or guidance could be needed.

This links to a limitation to the current use of CiTO annotation: the citation-level annotation may be
supported by some authoring systems (Markdown), even then the depiction may not be supported. If we convert
a Markdown file to a Microsoft Word file, the annotation is kept in the bibliography. However, when
writing a manuscript in Microsoft Word, LibreOffice, or Google Docs, combined with a reference manager
like Zotero, the CiTO annotation cannot be stored as part of the manuscript easily. The problem here is
that CiTO annotation in the reference manager are no longer linked to when they are cited. The workaround
is to add the CiTO annotation after completion of the Word document, directly to the bibliography.

LaTeX users may find them in a situation between that of Markdown users and reference manager users:
only if a manuscript-level Bib(La)TeX file is used the CiTO annotation can be added as notes to the
BiBTeX file. This is explained in this guidance document: https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/latex.html.

@@

With fourteen articles published in the CiTO Collection, the adoption is not as high as one would have hoped.
Nevertheless, the pilot is triggering interest with both authors and publishing platforms. The
*Journal of Cheminformatics* has already published a few more articles with CiTO annotation and
a search on the preprint servers ResearchSquare and ChemrXiv show a few more manuscripts. The support
by BioHackrXiv is a nice example of adoption beyond BioMedCentral.

The Scholia
use case shows how the information can be used downstream. Further uptake of this idea of typed citations
depends on the combined willingness of journal editors, authors, publishers and indexing services alike.
The rise of services like scite.ai shows that the research community needs this kind of information.

#### Availability of data and materials
CiTO annotation in the *Journal of Cheminformatics* are available from the journal's articles. CiTO
annotation data in Scholia is available from [Wikidata](https://wikidata.org/). Markdown templates that
support CiTO are available from https://github.com/jcheminform/markdown-jcheminf and
https://github.com/biohackrxiv/publication-template.

@@

#### Funding
Part of this work was supported by ELIXIR, the research infrastructure for life-science data.
